### 2014-10-07 paper/include/background.tex
spellchecked

DRAM is hierarichally organized as in \reffig{fig:dram_orga}. At the top of the hierarchy are the channels that are connected to one or more ranks, each composed of multiple storage arrays called banks. Each channel has individual address, data, and command buses that makes it possible for the channels to operate cuncurrently. All ranks on one channel can also operate in parallel constrained by the channel bandwith which is shared among the ranks. Further, the banks within the ranks are also able to operate in parallel, but are constrianed by both the shared channel bandwith and resources shared among the banks.

The DRAM cell array in each bank is structured as in \reffig{fig:dram_orgb}. One cell consists of a transistor and a capacitor, where the transistor is controlled by the wordline wire and connects the capacitor to the bitline wire. The wordline connects a number of cells which forms a row. Each bitline is connected to a sens amplifier and the row of sense amplifier constitutes the row buffer. Data is stored as a charge on the cell capacitor.

\begin{figure*}[t]
    \centering
    \subfigure[DRAM hierarcy]{
        \centering
        \includegraphics[width=0.4\textwidth]{dram_orga}
		\label{fig:dram_orga}
    }
    \subfigure[DRAM bank structure]{
        \centering
        \includegraphics[width=0.4\textwidth]{dram_orgb}
		\label{fig:dram_orgb}
    }
	\caption{DRAM system organization \cite{raidr}.}
	\label{fig:dram_org}
\end{figure*}

To read a row the wordline connected to the rows transistors must first be activated. When the wordline is activated, the transistors connects the capacitors to the bitlines. When the capacitors connects to the bitlines, charge will either flow from or to the capacitors depending on if the capacitors were charged or not, respectively. For this to work, the bitlines has to be precharged to \(V_{DD}\)/2. The sense amplifiers then detect the change of voltage, which makes it possible to interpret the stored data. The charge stored on the capacitors has, as mentioned, left the capacitor and a read operation is thus destructive. To maintain the data the capacitors are recarged by the bitlines, which are driven to either either \(V_{DD}\) or 0 V by the sense amplifiers depending on the former detected voltage change. Finally, if another row on the bank shall be accessed the open row has to be closed, which disconnects the capacitors from the bitlines.

To write to a row the corresponding wordline is first activated. As during a read, the transistors then connects the capacitors to the bitlines and charge will either flow from or to the capacitor. The row buffer are then loaded with the data that are driven to the bitlines by the sense amplifiers, thus storing the information on the capacitors. The row will be closed when another row shall be accessed.

@@

The memory controller (MC) normally issues the refresh operation periodically at an interval of 64 ms for each row, an interval that has been constant for the latests DRAM specifications [source]. This interval is halved if the DRAM exeeds the specified working temprature of 85$^{\circ}$C. %The average time between refresh commands are issued by the MC is denoted \(t_{REFI}\) and the time it takes to perform the refresh is denoted \(t_{RFC}\), i.e the time where no normal access requests can be serviced. % Maybe not needed, we shall see.

To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the 64ms interval. This approch is called Burst refresh and has some downsides. The approch blocks normal accesses for a longer period and form a peak of power consumption during the burst time. A better alternative is to distribute the refreshes across the interval, resulting in a lower average latency for the processor and a more stable power consumption.

@@

\begin{figure}[t]
    \centering
    \subfigure[DRAM throughput loss]{
        \centering
        \includegraphics[width=0.33\textwidth]{dram_throughput}
        \label{fig:dram_throughput}
    }
    \subfigure[DRAM power consumption]{
        \centering
        \includegraphics[width=0.33\textwidth]{dram_power}
        \label{fig:dram_power}
    }
    \caption{Projections on DRAM throughput loss and power consumption in extended-temprature operation \cite{raidr}.}
    \label{fig:dram_data_proj}
\end{figure}

The trends of DRAM chips is that they become larger and larger. In practice, this is realized to a high degree by adding additional rows in the banks. The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favourable as it increases the power consumption per row access.

As refresh operations are made by row, the refresh operations will increase liniary with the device capacity. In turn, the throughput loss increases exponentially as in \reffig{fig:dram_throughput}. Additional refresh operations gives larger power consumption and the refreshes becomes a growing portion of the total DRAM energy consumption as seen in \reffig{fig:dram_power}. When the device capacity reaches 64 GB the refresh power nearly constitute half of the total consumption, an increase from 15\% if the device capacity is 8 GB.

>>>>

DRAM is hierarchically organized as in \reffig{fig:dram_orga}. At the top of the hierarchy are the channels that are connected to one or more ranks, each composed of multiple storage arrays called banks. Each channel has individual address, data, and command buses that makes it possible for the channels to operate concurrently. All ranks on one channel can also operate in parallel constrained by the channel bandwidth which is shared among the ranks. Further, the banks within the ranks are also able to operate in parallel, but are constrained by both the shared channel bandwidth and resources shared among the banks.

The DRAM cell array in each bank is structured as in \reffig{fig:dram_orgb}. One cell consists of a transistor and a capacitor, where the transistor is controlled by the word line wire and connects the capacitor to the bit line wire. The word line connects a number of cells which forms a row. Each bit line is connected to a sens amplifier and the row of sense amplifier constitutes the row buffer. Data is stored as a charge on the cell capacitor.

\begin{figure*}[t]
    \centering
    \subfigure[DRAM hierarchy]{
        \centering
        \includegraphics[width=0.4\textwidth]{dram_orga}
		\label{fig:dram_orga}
    }
    \subfigure[DRAM bank structure]{
        \centering
        \includegraphics[width=0.4\textwidth]{dram_orgb}
		\label{fig:dram_orgb}
    }
	\caption{DRAM system organization \cite{raidr}.}
	\label{fig:dram_org}
\end{figure*}

To read a row the word line connected to the rows transistors must first be activated. When the word line is activated, the transistors connects the capacitors to the bit lines. When the capacitors connects to the bit lines, charge will either flow from or to the capacitors depending on if the capacitors were charged or not, respectively. For this to work, the bit lines has to be precharged to \(V_{DD}\)/2. The sense amplifiers then detect the change of voltage, which makes it possible to interpret the stored data. The charge stored on the capacitors has, as mentioned, left the capacitor and a read operation is thus destructive. To maintain the data the capacitors are recharged by the bit lines, which are driven to either either \(V_{DD}\) or 0 V by the sense amplifiers depending on the former detected voltage change. Finally, if another row on the bank shall be accessed the open row has to be closed, which disconnects the capacitors from the bit lines.

To write to a row the corresponding word line is first activated. As during a read, the transistors then connects the capacitors to the bit lines and charge will either flow from or to the capacitor. The row buffer are then loaded with the data that are driven to the bit lines by the sense amplifiers, thus storing the information on the capacitors. The row will be closed when another row shall be accessed.

@@

The memory controller (MC) normally issues the refresh operation periodically at an interval of 64 ms for each row, an interval that has been constant for the latest DRAM specifications [source]. This interval is halved if the DRAM exceeds the specified working temperature of 85$^{\circ}$C. %The average time between refresh commands are issued by the MC is denoted \(t_{REFI}\) and the time it takes to perform the refresh is denoted \(t_{RFC}\), i.e the time where no normal access requests can be serviced. % Maybe not needed, we shall see.

To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the 64ms interval. This approach is called Burst refresh and has some downsides. The approach blocks normal accesses for a longer period and form a peak of power consumption during the burst time. A better alternative is to distribute the refreshes across the interval, resulting in a lower average latency for the processor and a more stable power consumption.

@@

\begin{figure}[t]
    \centering
    \subfigure[DRAM throughput loss]{
        \centering
        \includegraphics[width=0.33\textwidth]{dram_throughput}
        \label{fig:dram_throughput}
    }
    \subfigure[DRAM power consumption]{
        \centering
        \includegraphics[width=0.33\textwidth]{dram_power}
        \label{fig:dram_power}
    }
    \caption{Projections on DRAM throughput loss and power consumption in extended-temperature operation \cite{raidr}.}
    \label{fig:dram_data_proj}
\end{figure}

The trends of DRAM chips is that they become larger and larger. In practice, this is realized to a high degree by adding additional rows in the banks. The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favorable as it increases the power consumption per row access.

As refresh operations are made by row, the refresh operations will increase linear with the device capacity. In turn, the throughput loss increases exponentially as in \reffig{fig:dram_throughput}. Additional refresh operations gives larger power consumption and the refreshes becomes a growing portion of the total DRAM energy consumption as seen in \reffig{fig:dram_power}. When the device capacity reaches 64 GB the refresh power nearly constitute half of the total consumption, an increase from 15\% if the device capacity is 8 GB.
